---
title: "Machine Learning Assignment 1"
author: "Agustín Valencia - aguva779"
date: "11/19/2019"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(openxlsx)
library(kknn)
library(ggplot2)
library(MASS)
```

# Assignment 1. Spam classification with nearest neighbors

```{r, echo=FALSE}
data <- read.xlsx("data/spambase.xlsx")
n = dim(data)[1]
set.seed(12345)
id = sample(1:n, floor(n*0.5))
train = data[id,]
test = data[-id,]
```

### 2. Use logistic regression to classify the training and test data by the classification principle $\hat{Y} = 1$ if $p(Y=1 | X ) > 0.5$, otherwise $\hat{Y} = 0$ and report the confusion matrices and the misclassification rates for train and test data. Analyze the obtained results. 


```{r, warning=FALSE, echo=FALSE}
# util function
get_performance <- function(targets, predictions, text) {
    cat("Classification Performance :", text, "\n")
    t <- table(targets, predictions)
    print("Confusion Matrix")
    print(t)
    tn <- t[1,1]
    tp <- t[2,2]
    fp <- t[1,2]
    fn <- t[2,1]
    total <- dim(test)[1]
    tpr <- tp/(tp+fp) * 100
    tnr <- tn/(tn+fn) * 100
    fpr <- fp/(tp+fp) * 100
    fnr <- fn/(tn+fn) * 100
    
    cat("Rates details:\n")
    cat(" TPR =", tpr, "% -")
    cat(" TNR =", tnr, "% -")
    cat(" FPR =", fpr, "% -")
    cat(" FNR =", fnr, "%")
    cat("\n Misclassification Rate = ", (fp+fn)/total * 100, "%\n")
}

# fit the model
fit <- glm(Spam ~ . , data = train, family = "binomial")
```

Evaluating the model with training data :

```{r, warning=FALSE, echo=FALSE}
# performance on training data
pred_train <- predict(fit, newdata = train)
pred_train_at_05 <- as.integer(pred_train > 0.5)
targets <- train$Spam
get_performance(targets, pred_train_at_05, "train set - trigger = 0.5")
```

Now, with unseen data it can be observed that the misclassification rate increased, though numbers still consistent.

```{r, warning=FALSE, echo=FALSE}
# performance on test data
pred_test <- predict(fit, newdata = test)
pred_test_at_05  <- as.integer(pred_test > 0.5)
targets <- test$Spam
get_performance(targets, pred_test_at_05, "test set - trigger = 0.5")
```


### 3. Use logistic regression to classify the test data by the classification principle  $\hat{Y} = 1$ if $p(Y=1 | X ) > 0.8$, otherwise $\hat{Y} = 0$ 

Setting a higher trigger implies that the classifier will be more selective, then it is expected to decrease the amount of mails being labeled as spam. 

The training stats:

```{r, echo=FALSE}
# performance on train data
pred_train_at_08  <- as.integer(pred_train > 0.8)
targets <- train$Spam
get_performance(targets, pred_train_at_08, "train set - trigger = 0.8")
```

Testing stats: 

```{r, echo=FALSE}
# performance on test data
pred_test_at_08  <- as.integer(pred_test > 0.8)
targets <- test$Spam
get_performance(targets, pred_test_at_08, "test set - trigger = 0.8")
```

Although the misclassification rate has increased, the false positive rate, i.e., the amount of valid email being sent to the spambox, has decreased, which from a user perspective could be more valuable than a higher accuracy on true positives. 

### 4. Use standard kknn() with K = 30 from package *kknn*, report the misclassification rates for the training and test data and compare the results with step 2.

```{r, warning=FALSE, echo=FALSE}
# Train KNN K=30
knn_model <- train.kknn(Spam ~ . , data = train, ks = 30)

# performance on training data
knn_fit <- predict(knn_model, train)
results <- as.integer(knn_fit > 0.5)
target <- train$Spam
get_performance(target, results, "train knn - k = 30")

# performance on test data
knn_fit <- predict(knn_model, test)
results <- as.integer(knn_fit > 0.5)
target <- test$Spam
get_performance(target, results, "test knn - k = 30")
```


### 5. Repeat step 4 for K=1 and compare results with step 4. What effects does the decrease of K lead to and why?

```{r, warning=FALSE, echo=FALSE}
# Train KNN K=1
knn_model <- train.kknn(Spam ~ . , data = train, ks = 1)

# performance on training data
knn_fit <- predict(knn_model, train)
results <- as.integer(knn_fit > 0.5)
target <- train$Spam
get_performance(target, results, "train knn - k = 1")

# performance on test data
knn_fit <- predict(knn_model, test)
results <- as.integer(knn_fit > 0.5)
target <- test$Spam
get_performance(target, results, "test knn - k = 1")
```

If we assign k=1 training misclassification is 0%, this means we are overfitting our model, thus the misclassification for the testing set may be bigger than other scenarios. 






# Assignment 3. Feature selection by cross-validation in a linear model. 





















# Assignment 4. Linear regression and regularization

### 1. Import data and create a plot of Moisture versus Protein. Do you think these data are described well by a linear model? 

```{r, echo=FALSE}
## Import data and plot Moisture vs Protein.
data <- read.xlsx("data/tecator.xlsx")
plot(data$Protein, data$Moisture)
```

By the plot, although there are some outliers, it seems that the data could be approximated by a linear model. 

### 2. Consider model $M_i$ in which Moisture is normally distributed and the expected Moisture is polynomial 

*PUT DOWN SOME MATH BAAAAAM !!*

### 3. Divide the data (50/50) and fit models $M_i, i=1,\cdots, 6$. For each model, record the training and validation MSE and present a plot showing how training and validation MSE depend on $i$. Which model is best according to this plot? How do MSE values change and why? Interpret this picture in bias-variance tradeoff.

```{r, echo=FALSE}
# Creating data sets
p <- data$Protein
y <- data$Moisture
P <- data.frame(
    Y = y,
    P1 = p, 
    P2 = p^2, 
    P3 = p^3, 
    P4 = p^4, 
    P5 = p^5, 
    P6 = p^6 
)

n = dim(P)[1]
set.seed(12345)
id = sample(1:n, floor(n*0.5))
train = P[id,]
test = P[-id,]

# Training models
M1 <- lm(Y ~ ., data = train[,1:2])
M2 <- lm(Y ~ ., data = train[,1:3])
M3 <- lm(Y ~ ., data = train[,1:4])
M4 <- lm(Y ~ ., data = train[,1:5])
M5 <- lm(Y ~ ., data = train[,1:6])
M6 <- lm(Y ~ ., data = train[,1:7])

## Train Scores
eval_model <- function(model, data) {
  pred <- predict(model, data)
  errors <- pred - data$Y
  MSE <- mean(errors^2)
  return(MSE)
}

MSE_train <- c()
MSE_train[1] <- eval_model(M1, train)
MSE_train[2] <- eval_model(M2, train)
MSE_train[3] <- eval_model(M3, train)
MSE_train[4] <- eval_model(M4, train)
MSE_train[5] <- eval_model(M5, train)
MSE_train[6] <- eval_model(M6, train)

MSE_test <- c()
MSE_test[1] <- eval_model(M1, test)
MSE_test[2] <- eval_model(M2, test)
MSE_test[3] <- eval_model(M3, test)
MSE_test[4] <- eval_model(M4, test)
MSE_test[5] <- eval_model(M5, test)
MSE_test[6] <- eval_model(M6, test)

df <- data.frame(
    degree <- c(1:6),
    MSE_train,
    MSE_test
)
p <- ggplot()
p <- p +geom_point(data = df, aes( x = degree, y = MSE_train, color="Train")) + 
    geom_line(data = df, aes( x = degree, y = MSE_train, color="Train"))
p <- p + geom_point(data = df, aes( x = degree, y = MSE_test, color="Test")) + 
    geom_line(data = df, aes( x = degree, y = MSE_test, color="Test"))
p <- p + labs(y="MSE", colour="Dataset", title = "Mean Square Errors") + geom_line() 
p

```

### 4. Perform variable selection of a linear model in which Fat is response and Channel1-Channel100 are predictors by using stepAIC. Comment on how many variables were selected.

```{r, echo=FALSE, results="hide"}
adhok_data <- data[,2:101]
Fat <- data$Fat
adhok_data <- cbind(adhok_data, Fat)
model <- lm(Fat ~ . , data=adhok_data) 
step <- stepAIC(model, direction ="both")
selected_vars <- step$coefficients
```
```{r}
cat("There were selected", length(selected_vars), "variables\n")
```































 