---
title: "Machine Learning Assignment 1"
author: "Agustín Valencia"
date: "11/19/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(openxlsx)
library(kknn)
```

# Assignment 1. Spam classification with nearest neighbors

### 1. Importing the data: 

```{r}
data <- read.xlsx("data/spambase.xlsx")
n = dim(data)[1]
set.seed(12345)
id = sample(1:n, floor(n*0.5))
train = data[id,]
test = data[-id,]
```

### 2. Use logistic regression to classify the training and test data by the classification principle $\hat{Y} = 1$ if $p(Y=1 | X ) > 0.5$, otherwise $\hat{Y} = 0$ and report the confusion matrices and the misclassification rates for train and test data. Analyze the obtained results. 


```{r, warning=FALSE}
# util function
get_performance <- function(targets, predictions) {
    t <- table(targets, predictions)
    print("Confusion Matrix")
    print(t)
    tn <- t[1,1]
    tp <- t[2,2]
    fp <- t[1,2]
    fn <- t[2,1]
    total <- dim(test)[1]
    tpr <- tp/total * 100
    tnr <- tn/total * 100
    fpr <- fp/total * 100
    fnr <- fn/total * 100
    
    cat("Classification performance:\n")
    cat("TPR = ", tpr, "%\n")
    cat("TNR = ", tnr, "%\n")
    cat("FPR = ", fpr, "%\n")
    cat("FNR = ", fnr, "%\n")
    cat("Misclassification Rate = ", (fp+fn)/total * 100, "%\n")
}

# fit the model
fit <- glm(Spam ~ . , data = train, family = "binomial")

# performance on training data
pred_train <- predict(fit, newdata = train)
pred_train_at_05 <- as.integer(pred_train > 0.5)
targets <- train$Spam
get_performance(targets, pred_train_at_05)

# performance on test data
pred_test <- predict(fit, newdata = test)
pred_test_at_05  <- as.integer(pred_test > 0.5)
targets <- test$Spam
get_performance(targets, pred_test_at_05)
```





### 3. Use logistic regression to classify the test data by the classification principle  $\hat{Y} = 1$ if $p(Y=1 | X ) > 0.8$, otherwise $\hat{Y} = 0$ 

```{r}
# performance on train data
pred_train_at_08  <- as.integer(pred_train > 0.8)
get_performance(targets, pred_train_at_08)

# performance on test data
pred_test_at_08  <- as.integer(pred_test > 0.8)
get_performance(targets, pred_test_at_08)
```

### 4. Use standard kknn() with K = 30 from package *kknn*, report the misclassification rates for the training and test data and compare the results with step 2.

```{r, warning=FALSE}
# Train KNN K=30
knn_model <- train.kknn(Spam ~ . , data = train, ks = 30)

# performance on training data
knn_fit <- predict(knn_model, train)
results <- as.integer(knn_fit > 0.5)
target <- train$Spam
get_performance(target, results)

# performance on test data
knn_fit <- predict(knn_model, test)
results <- as.integer(knn_fit > 0.5)
target <- test$Spam
get_performance(target, results)
```


### 5. Repeat step 4 for K=1 and compare results with step 4. What effects does the decrease of K lead to and why?

```{r, warning=FALSE}
# Train KNN K=1
knn_model <- train.kknn(Spam ~ . , data = train, ks = 1)

# performance on training data
knn_fit <- predict(knn_model, train)
results <- as.integer(knn_fit > 0.5)
target <- train$Spam
get_performance(target, results)

# performance on test data
knn_fit <- predict(knn_model, test)
results <- as.integer(knn_fit > 0.5)
target <- test$Spam
get_performance(target, results)
```

If we assign k=1 training misclassification is 0%, this means we are overfitting our model, thus the misclassification for the testing set may be bigger than other scenarios. 







































 